Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-07 analysis/paper/paper.Rmd
work on discussion

The variation of pottery shape for each phase is explored by the distributions of the first three PC scores to compare the extent of standardization across phases. Figure \@ref(fig:PCs-distribution) suggests variations of PC scores for each vessel by phase. The first PC, the hight of the vessels, shows more variations in the pre-European period compared to the Chinese period. That indicates standardization appears strong in the Chinese period but weak in the pre-European period. The second PC also presents a similar pattern that stronger standardization of shape in the Chinese period compared to the other two phases. To see whether the differences in the distribution of PCs between any two phases is statistically significant, coefficients of variation statistic (CVs) among phase groups of PCs is examined with a modified signed-likelihood ratio test for the equality of CVs. P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value < 0.01) in the standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European Figure \@ref(fig:CV-test-PCs). Significant difference is also detected between pre-Eruopean and post-European (P-value < 0.05).

@@

@@

The results confirm that the differences in pottery shapes can be detected using oEFA-based outline analysis approach. There are differences in pottery shapes detected when indigenous society encountered the foreign presence, including the European in the 17th century and the Chinese in the 19th century. The significant difference especially presented between Chinese contact and either pre-European or post-European (P-value < 0.01). Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments, during the presence of the European that hints the emergence of social inequality. We further examine the ceramic production that usually used as an indicator as the presence of social differentiation, such as the elite control [@Costin2001; @Junker1999]. Although the difference was detected before and after the European presence, the pottery during the Chinese period show more standardized shape in general. This might indicate the appearance of specialized group. Given the decline of the indigenous population in the Chinese period might be a cause of a more standardized shape [@Chen2007], we examined the spatial distribution of pottery selected in this paper. The pottery distributed across the sampling area without any clusters could rule out the bias caused by the population dynamics.

The nature and context of the interaction with the foreign presences in indigenous society might give some insights into the different degree of standardized pottery among those contact phases.

It is interesting to find the significant difference is observed during the Chinese period that suggests a more standardized pottery shape compared with the previous two periods.

When it comes to ceramic specialization, metric measurements and clay paste are two major indicators. Shape analysis could be viewed as a replacement of metric measurement that enables subtle changes to be detected. We observe the increasing standardization of the pottery shape, however, it is interesting to note that there is no obvious changes in the clay paste across different phases. More standardized shape without changes in raw materials indicate a

Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels? How does pottery change in shape throughout 500 years from a single site that can represent the development of social organization in northeastern Taiwan? Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants?

>>>>

The variation of pottery shape for each phase is explored by the distributions of the first three PC scores to compare the extent of standardization across phases. Figure \@ref(fig:PCs-distribution) suggests variations of PC scores for each vessel by phase. The first PC, the hight of the vessels, shows more variations in the pre-European period compared to the Chinese period. That indicates standardization appears strong in the Chinese period but weak in the pre-European period. The second PC also presents a similar pattern that stronger standardization of shape in the Chinese period compared to the other two phases. To see whether the differences in the distribution of PCs between any two phases is statistically significant, coefficients of variation statistic (CVs) among phase groups of PCs is examined with a modified signed-likelihood ratio test for the equality of CVs. P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value < 0.01) in the standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European Figure \@ref(fig:CV-test-PCs). Significant difference is also detected between pre-Eruopean and post-European (P-value < 0.05). Given the decline of the indigenous population in the Chinese period might be a cause of a more standardized shape [@Chen2007], we examined the spatial distribution of pottery selected in this paper. The pottery distributed across the sampling area without any clusters could rule out the bias caused by the population dynamics.

@@

1. Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels? 2. How does pottery change in shape throughout 500 years from a single site that can represent the development of social organization in northeastern Taiwan? 
3. Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants?

@@

Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments, during the presence of the European that hints the emergence of social inequality. Using an EFA-based outline approach, we examine the degree of the ceramic standardization as an indicator of the presence of ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999]. The results confirm that the differences in pottery shapes can be detected using EFA method. There are differences in pottery shapes detected in the indigenous society across different foreign contact phases, including the European in the 17th century and the Chinese in the 19th century. The significant difference especially presented between Chinese contact and either pre-European or post-European (P-value < 0.01). We also observe there is no obvious changes in the clay paste across different phases in relation to the increasing standardization of the pottery shape. More homogeneous and standardized in shape without changes in raw materials reflects the clay source might be the same but a preferred shape was emphasized during the Chinese contact period. This might indicate the appearance of specialized group. The pottery shape during the Chinese period tend to be short with oval-shaped body showing wider body diameter compared to previous two phases. It would need further analysis to understand if the changes in shape related to the function of pot.

The nature and context of the interaction with the foreign presences in indigenous society might give an insight into the different degree of standardized pottery among those contact phases.
